---
title: "QSP Homework 1"
author: "Anjali Venna"
date: "2025-02-07"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


# Part 1: Implement Model for O2 Homeostasis at Initial Conditions

At initial conditions, the O2 homeostasis model shows a rate of change of zero for RBC, EPO, and O2.

```{r Part1}
# Libraries used
library(deSolve)
library(ggplot2)
library(dplyr)
library(knitr)


# Define the system of ODEs
model <- function(t, y, params) {
  with(as.list(c(y, params)), {
    dRBC <- a1 * SC * (1 + 0.5 * EPO^4 / (2^4 + EPO^4)) - b1 * RBC
    dO2  <- a2 * RBC - b2 * O2
    dEPO <- a3 * O2^g3 - b3 * EPO
    return(list(c(dRBC, dO2, dEPO)))
  })
}

# Define parameters
params <- c(a1 = 0.1, b1 = 0.15, a2 = 0.1, b2 = 1.0, a3 = 100, 
            b3 = 0.1, g3 = -1, SC = 1000)

# Initial conditions for RBC, O2, and EPO
y0 <- c(RBC = 1000, O2 = 100, EPO = 10)

# Time sequence for simulation
times <- seq(0, 70, by = 0.1)

# Solve the system of ODEs
result_1 <- ode(y = y0, times = times, func = model, parms = params)

# Convert 'result' (matrix) to a dataframe for ggplot2
df <- as.data.frame(result_1)
df_long <- reshape2::melt(df, id.vars = "time")

# Plot using ggplot2
ggplot(df_long, aes(x = time, y = value, color = variable)) +
  geom_line(size = 1) +
  labs(x = "Time", y = "Concentration", color = "Legend", title = "O2 Homeostasis Model (Initial Conditions)\n RBC = 1000 | O2 = 100 | EPO = 10") +
  theme_minimal() +
  theme(legend.position = "right", plot.title = element_text(hjust = 0.5)) +
      scale_color_manual(values=c('Red','Blue', "Green"))

# Creating a data table
# Assuming you already have a dataset with a time column
# Example: result is your data frame with a 'time' column
# df1 <- df %>%
#   mutate(formatted_time = sapply(time, function(t) {
#     h <- floor(t)  # Extract whole hours
#     m <- round((t - h) * 60)  # Convert fractional part to minutes
#     paste(h, "hrs", m, "mins")
#   }))
# 
# kable(as.data.frame(df1), digits = 5, caption = "ODE Simulation Results (Steady State)")
```

# Part 2: Perform other simulations and document results

Part 2a: Change initial conditions, one at a time

Decreasing Initial Conditions

```{r Part2a, fig.show='hold'}
# Load necessary libraries
library(deSolve)  # For ODE solving
library(ggplot2)  # For plotting
library(reshape2) # For melting data

# List of initial conditions to test (changing RBC, O2, and EPO values)
initial_conditions_list <- list(
  c(RBC = 700, O2 = 100, EPO = 10),
  c(RBC = 1000, O2 = 50, EPO = 10),
  c(RBC = 1000, O2 = 100, EPO = 1)
)

# Loop through each initial condition
for (i in seq_along(initial_conditions_list)) {
  
  # Extract current initial conditions
  initial_conditions <- initial_conditions_list[[i]]
  
  # Solve the ODE system
  result_2 <- ode(y = initial_conditions, times = times, func = model, parms = params)
  
  # Convert 'result' (matrix) to a dataframe for ggplot2
  df_2 <- as.data.frame(result_2)
  df_long_2 <- reshape2::melt(df_2, id.vars = "time")
  
  # Create a dynamic title based on the current initial conditions
  plot_title <- paste("O2 Homeostasis Model (Changed Conditions)",  
                      "\nRBC =", initial_conditions["RBC"], 
                      "| O2 =", initial_conditions["O2"], 
                      "| EPO =", initial_conditions["EPO"])
  
  # Create plot
  p <- ggplot(df_long_2, aes(x = time, y = value, color = variable)) +
    geom_line(size = 1) +
    labs(x = "Time", y = "Concentration", color = "Legend", title = plot_title) +
    theme_minimal() +
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5)) + 
    scale_color_manual(values=c('Red','Blue', "Green"))
  
    print(p)  # Explicitly print the plot inside the loop
  # 
   df2 <- df_2 %>%
   mutate(formatted_time = sapply(time, function(t) {
     h <- floor(t)  # Extract whole hours
     m <- round((t - h) * 60)  # Convert fractional part to minutes
     paste(h, "hrs", m, "mins")
   }))
   
   print(kable(as.data.frame(df2), digits = 5, caption = plot_title))

}

```

Increasing Initial Conditions
```{r}

```

Part 2b: Change parameters values, one at a time

Decreasing

```{r Part2b}
# List of changed parameters to test
changed_param_list <- list(
  c(a1 = 0.05, b1 = 0.15, a2 = 0.1, b2 = 1.0, a3 = 100, b3 = 0.1, g3 = -1, SC = 1000),
  c(a1 = 0.1, b1 = 0.07, a2 = 0.1, b2 = 1.0, a3 = 100, b3 = 0.1, g3 = -1, SC = 1000),
  c(a1 = 0.1, b1 = 0.15, a2 = 0.05, b2 = 1.0, a3 = 100, b3 = 0.1, g3 = -1, SC = 1000),
  c(a1 = 0.1, b1 = 0.15, a2 = 0.1, b2 = 0.5, a3 = 100, b3 = 0.1, g3 = -1, SC = 1000),
  c(a1 = 0.1, b1 = 0.15, a2 = 0.1, b2 = 1.0, a3 = 50, b3 = 0.1, g3 = -1, SC = 1000),
  c(a1 = 0.1, b1 = 0.15, a2 = 0.1, b2 = 1.0, a3 = 100, b3 = 0.05, g3 = -1, SC = 1000),
  c(a1 = 0.1, b1 = 0.15, a2 = 0.1, b2 = 1.0, a3 = 100, b3 = 0.1, g3 = -1.5, SC = 1000),
  c(a1 = 0.1, b1 = 0.15, a2 = 0.1, b2 = 1.0, a3 = 100, b3 = 0.1, g3 = -1, SC = 500)
)

# Loop through each initial condition
for (i in seq_along(changed_param_list)) {
  
  # Extract current initial conditions
  changed_param <- changed_param_list[[i]]
  
  param_names <- c("a1", "b1", "a2", "b2", "a3", "b3", "g3", "SC")
  
  # Solve the ODE system
  result_3 <- ode(y = y0, times = times, func = model, parms = changed_param)
  
  # Convert 'result' (matrix) to a dataframe for ggplot2
  df_3 <- as.data.frame(result_3)
  df_long_3 <- reshape2::melt(df_3, id.vars = "time")
  
  # Create a dynamic title based on the current initial conditions
  plot_title <- paste("O2 Homeostasis Model - Case", i, "\nChanged Parameter =", param_names[i], "(", changed_param[i],")")
  
  # Create plot
  p <- ggplot(df_long_3, aes(x = time, y = value, color = variable)) +
    geom_line(size = 1) +
    labs(x = "Time", y = "Concentration", color = "Legend", title = plot_title) +
    theme_minimal() +
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5))
  
    print(p)  # Explicitly print the plot inside the loop
  # 
  # df2 <- df_2 %>%
  # mutate(formatted_time = sapply(time, function(t) {
  #   h <- floor(t)  # Extract whole hours
  #   m <- round((t - h) * 60)  # Convert fractional part to minutes
  #   paste(h, "hrs", m, "mins")
  # }))
  # 
  # print(kable(as.data.frame(df2), digits = 5, caption = plot_title))

}

```


Part 2b: Study Combinations of Changes

1.	What if lower g3 but higher SC?
2.	What if lower A2, higher RBC?
3.	What if lower A2, lower O2?

```{r Part 2b}

# Lower g3, Higher SC

  # Define parameters
  params <- c(a1 = 0.1, b1 = 0.15, a2 = 0.1, b2 = 1.0, a3 = 100, 
              b3 = 0.1, g3 = -1.5, SC = 1500)
  
  # Initial conditions for RBC, O2, and EPO
  y0 <- c(RBC = 1000, O2 = 100, EPO = 10)
  
  # Time sequence for simulation
  times <- seq(0, 70, by = 0.1)
  
  # Solve the system of ODEs
  result_1 <- ode(y = y0, times = times, func = model, parms = params)
  
  # Convert 'result' (matrix) to a dataframe for ggplot2
  df <- as.data.frame(result_1)
  df_long <- reshape2::melt(df, id.vars = "time")
  
  # Plot using ggplot2
  ggplot(df_long, aes(x = time, y = value, color = variable)) +
    geom_line(size = 1) +
    labs(x = "Time", y = "Concentration", color = "Legend", title = "Lower g3 [-1.5], Higher SC [1500]") +
    theme_minimal() +
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5)) +
        scale_color_manual(values=c('Red','Blue', "Green"))


# Lower A2, Higher RBC
  params <- c(a1 = 0.1, b1 = 0.15, a2 = 0.05, b2 = 1.0, a3 = 100, 
              b3 = 0.1, g3 = -1, SC = 1000)
  
  # Initial conditions for RBC, O2, and EPO
  y0 <- c(RBC = 1500, O2 = 100, EPO = 10)
  
  # Time sequence for simulation
  times <- seq(0, 70, by = 0.1)
  
  # Solve the system of ODEs
  result_1 <- ode(y = y0, times = times, func = model, parms = params)
  
  # Convert 'result' (matrix) to a dataframe for ggplot2
  df <- as.data.frame(result_1)
  df_long <- reshape2::melt(df, id.vars = "time")
  
  # Plot using ggplot2
  ggplot(df_long, aes(x = time, y = value, color = variable)) +
    geom_line(size = 1) +
    labs(x = "Time", y = "Concentration", color = "Legend", title = "Lower a2 [0.05], Higher RBC [1500]") +
    theme_minimal() +
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5)) +
        scale_color_manual(values=c('Red','Blue', "Green"))


# Lower A2, Lower O2
  # Define parameters
  params <- c(a1 = 0.1, b1 = 0.15, a2 = 0.05, b2 = 1.0, a3 = 100, 
              b3 = 0.1, g3 = -1, SC = 1000)
  
  # Initial conditions for RBC, O2, and EPO
  y0 <- c(RBC = 1000, O2 = 25, EPO = 10)
  
  # Time sequence for simulation
  times <- seq(0, 70, by = 0.1)
  
  # Solve the system of ODEs
  result_1 <- ode(y = y0, times = times, func = model, parms = params)
  
  # Convert 'result' (matrix) to a dataframe for ggplot2
  df <- as.data.frame(result_1)
  df_long <- reshape2::melt(df, id.vars = "time")
  
  # Plot using ggplot2
  ggplot(df_long, aes(x = time, y = value, color = variable)) +
    geom_line(size = 1) +
    labs(x = "Time", y = "Concentration", color = "Legend", title = "Lower a2 [0.05], Lower O2 [25]") +
    theme_minimal() +
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5)) +
        scale_color_manual(values=c('Red','Blue', "Green"))

```




